The role of aneuploidy in the evolution of cancer drug resistance

### 2023-12-04 EvolutionaryRescue1.tex
edited abstract

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes and observe that aneuploidy plays an important role in the survival of the cancer cell population when the initial tumor size is not too large and the effect of the anti-cancer drug on the aneuploid cells is not too severe.
\end{abstract}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. Chromosomal instability (aneuploidy) occurs frequently in cancer cells and introduces large genetics, suggesting that it may be an important mechanism by which populations of cancer cells to adapt to chemotherapy. To determine when this should be the case, we model a population of cancer cells exposed to chemotherapy that can achieve evolutionary rescue with or without aneuploidy. We use branching process approximations to obtain simple expressions for the probability of evolutionary rescue and the effect of aneuploidy. We find that even when aneuploidy by itself only confers partial drug resistance, it can still be an important stepping stone on the way to evolutionary rescue, allowing even small tumors to produce enough genetic diversity to avoid extinction.
\end{abstract}
